Finiteness, Cardinality, and Combinatorics in Homotopy Type Theory

### 2020-01-06 paper.tex
much shorter

  We formalise our work in Cubical Agda
  \cite{vezzosiCubicalAgdaDependently2019}, and we implement a library for proof
  search (including combinators for level-polymorphic fully generic currying),
  and demonstrate how it can be used to both prove properties and synthesise
  full functions given desired properties.
\end{abstract}
% \section{Introduction}
% Proofs in a constructive setting are more substantial things than their
% classical counterparts.
% Rather than just evidence for some fact, they are objects in their own right,
% which we can manipulate and extract information from.

% Sometimes, the extra detail can reveal interesting subtleties: fissures which
% divide homogeneous-seeming definitions into a variety of related concepts.
% Finiteness is a prime example of this phenomenon: in classical set theory, a
% finite set is a finite set, and there's not much else that we can say.
% In constructive type theory, we start the bidding at no fewer than \emph{four}
% different predicates representing finiteness
% \cite{spiwackConstructivelyFinite2010}, upped to eight in
% \cite{firsovVariationsNoetherianness2016a}: all of which differ in significant
% and interesting ways.

% Other times, the extra baggage is unwanted.
% Maybe we want to enforce that a particular piece of information should be
% hidden, or treated as irrelevant: ``yes, I have proven that there is a
% particular number with this property, but I don't want to reveal \emph{which}
% number.''

% This is where Homotopy Type Theory \cite{hottbook} can step in.
% Through primarily a generalisation of equality, we gain access to a number of
% exotic types which can, among other things, allow us to perform this ``hiding''.
% In the context of finiteness, the act of hiding is worth studying in its own
% right, and further delineates several new finiteness predicates which collapse
% even in standard Martin-Löf type theory
% \cite{martin-lofIntuitionisticTypeTheory1980}.

% The other headline feature of HoTT's generalised equality is univalence: put
% simply, it allows us to treat isomorphic types as equal, giving us in the
% constructive world access to a technique that is commonplace classically.
% It turns out that finite types have a lot to say about isomorphisms, and we will
% rely heavily on this aspect of HoTT to work with them.
% \subsection{Contributions}
% In this work, we will explore finite types in Cubical Type Theory
% \cite{cohenCubicalTypeTheory2016}, and expose their relationship to infinite
% types, species, and demonstrate their practical uses for proofs in dependently
% typed programming languages.
% \subsubsection{Strong Finiteness Predicates}
% We will first explore the ``strong'' notions of finiteness (i.e. those at least
% as strong as Kuratowski finiteness \cite{kuratowskiNotionEnsembleFini1920}),
% with a special focus on cardinal finiteness (section~\ref{cardinal}), and
% manifest enumerability (section~\ref{manifest-enumerability}), which is new, to
% our knowledge.

% \input{figures/finite-classification}
% Figure~\ref{finite-classification} organises the predicates according to
% their ``strength''; i.e. how much information they provide about a conforming
% type.
% For instance, a proof that some type \(A\) is manifestly Bishop finite (the
% strongest of the notions, explored in section~\ref{manifest-bishop}) also tells
% us that \(A\) is discrete (has decidable equality), and gives us a linear order
% on the type.
% A type that is Kuratowski finite (section~\ref{kuratowski}) has no such extra
% features: indeed, we will see examples of Kuratowski finite types which are not
% even sets, never mind discrete ones.

% We will go through each of the predicates, proving how to weaken each (i.e. we
% will provide a proof that every cardinally finite type is Kuratowski finite),
% and how to strengthen them, given the required property.
% In terms of figure~\ref{finite-classification}, this amounts to providing proofs
% for each arrow.

% We will---through the use of containers
% \cite{abbottContainersConstructingStrictly2005}---formally prove the equivalence
% these predicates have with the usual function relations i.e. we will show that a
% proof of manifest enumerability is precisely equivalent to a surjection from a
% finite prefix of the natural numbers.

% For each predicate, we will also prove its closure over sums and products in
% both dependent and non-dependent forms, if such a closure exists.
% This will culminate in our main result for this section: the formal proof that
% decidable Kuratowski finite sets form a topos. \todo[noline]{reference for this}
% More specifically, we show that cardinal finite sets form a topos, that
% decidable Kuratowski finite sets are equivalent in strength to cardinal finite
% sets, and carry the proof over.
% This result relies on proofs on each of the other finiteness predicates.
% \subsubsection{Species}
% \todo[inline]{Redo this next paragraph}
% In section~\ref{species-and-containers}, we will redefine our finiteness
% predicates using \emph{containers}, and prove that the new definitions are
% equivalent to the old.
% Using this, we will prove much stronger relationships between our finiteness
% predicates and relation-based definitions of finiteness.
% Finally, We will show how the species-container duality falls out naturally from
% these new definitions.

% \subsubsection{Infinite Types}
% In section~\ref{infinite-cardinalities}, we will extend our study of finite
% types to infinite but countable types.
% We will see that the finiteness predicates are mirrored with countable
% counterparts, and we will prove closure under the Kleene star and plus.

% \subsubsection{Practical Uses of Finiteness}
% Proofs of finiteness have well-known practical applications in
% constructive mathematics \cite{firsovDependentlyTypedProgramming2015}.
% In section~\ref{practical}, we build a library which exploits these uses in
% Cubical Agda \cite{vezzosiCubicalAgdaDependently2019}, allowing automation of
% complex proofs over finite types.
% We frame this in terms of the principle of omniscience for finite types.
% Thanks to the flexibility afforded to us by Cubical Type Theory, we are able to
% go further than the usual examples of this kind of proof automation: as well as
% proving properties about functions, we can synthesise functions whole-cloth from
% their desired properties.
% Through the unified interface for finite and countable types, we can
% reuse the automation machinery for \emph{partial} proof search over infinite
% search spaces.
% Along the way, we extend the work in \cite{allaisGenericLevelPolymorphic2019} to
% prove isomorphisms between the curried and uncurried forms of \(n\)-ary
% dependent functions.
\section{Split Enumerability} \label{split-enumerable}
We will start with the simplest definition of finiteness: we say a set is
enumerable if there is a list of its elements which contains every element in
the set.
More formally:
\begin{rm-definition}[Split Enumerable Set]
  \begin{equation}
    \mathcal{E}!(A) = \Sigma {(\mathit{xs} : \textbf{List}(A))} , \Pi {(x : A)} , x \in xs
  \end{equation}
  We call the first component of this pair the ``support'' list, and the second
  component the ``cover'' proof.
\end{rm-definition}

@@

  It suffices now to prove the following:
  \begin{equation}
    \Pi {(x : A)} , \Pi {(\mathit{xs} : \textbf{List}(A))} , x \in \mathit{xs} \rightarrow x \inunique \text{uniques}(\mathit{xs})
  \end{equation}
  And from that we can generate our cover proof.
\end{proof}
Note that while manifest Bishop finiteness as split enumerability are equivalent
in ``strength'', the two types are not equivalent.
In particular, there are infinitely many inhabitants of \(\mathcal{E}!\), while
for a type \(A\) with \(n\) inhabitants, there are only \(n!\) inhabitants of
\(\mathcal{B}(A)\).
\subsection{Closure}
Proving equal strength of split enumerability and manifest Bishop finiteness
allows us to carry all of the previous proofs of closure over to manifest Bishop
finite sets (and vice-versa).
Missing from our previous proofs was a proof of closure of functions.
We remedy that here.
\begin{rm-theorem}
  Manifest bishop finiteness is closed over dependent functions (\(\prod\)-types).

  Formally, given a type \(A\) and a type family \(U\) on \(A\), when \(A\) is
  manifestly Bishop finite:
  \begin{equation}
    \mathcal{B}\left( A \right)
  \end{equation}
  And \(U\) is manifestly Bishop finite over all points of \(A\):
  \begin{equation}
    \Pi {(x : A)} , \mathcal{B}\left( U(x) \right)
  \end{equation}
  Then we have the following:
  \begin{equation}
    \mathcal{B}\left(\Pi {(x : A)} , U(x)\right)
  \end{equation}
\end{rm-theorem}
\begin{proof}
  This proof is essentially the composition of two transport operations, made
  available to us via univalence.

@@

  We then prove an isomorphism between this representation and \(\Pi\)-types.
  \begin{equation}
    \mathbf{Tuple}(T, n) \iff \Pi {(x : \mathbf{Fin}(n))} , T(x)
  \end{equation}
  This allows us to transport our proof of finiteness on tuples to one on
  functions from \(\mathbf{Fin}\) (our second transport operation), proving our
  goal.

  \ExecuteMetaData[agda/Cardinality/Finite/ManifestBishop.tex]{pi-clos}
\end{proof}
\section{Manifest Enumerability} \label{manifest-enumerability}
A defining feature of split enumerability and Bishop finiteness is decidability:
both predicates imply a decidable equality function on the underlying type.
To find a predicate for finiteness which doesn't imply this, and therefore
works with types other than just sets, we \emph{truncate} the membership proof.
\begin{rm-definition}[Manifest Enumerability]
  \begin{equation}
    \mathcal{E}(A) = \Sigma {(\mathit{xs} : \mathbf{List}(A))} , \Pi {(x : A)} , \lVert x \in \mathit{xs} \rVert
  \end{equation}
\end{rm-definition}
\begin{rm-definition}[Propositional Truncation]a \label{prop-trunc}
  The type \(\lVert A \rVert\) on some type \(A\) is a propositionally truncated
  proof of \(A\).
  In other words, it is a proof that some \(A\) exists, but it does not tell you
  \emph{which} \(A\).

@@

  We will use and consume values of the type \(\lVert A \rVert\) in three main
  ways.
  \begin{enumerate}
    \item \label{elim-prop-prop} We can first eliminate from \(\lVert A \rVert\)
      into any type which is a proposition.
      In other words, given a function \(f : A \rightarrow B\), and a proof that
      \(B\) is a proposition, we can construct a function \(\lVert A \rVert
      \rightarrow B\).
    \item \label{elim-prop-monad} As a consequence of this first point, we can
      always eliminate into another propositionally truncated type.
      As a result, \(\lVert \cdot \rVert\) forms a Monad: for our purposes, this
      simply means tat we can work ``under'' a propositional truncation in an
      ergonomic way.
    \item \label{elim-prop-coh} We can eliminate from \(\lVert A \rVert\) with a
      function \(f : A \rightarrow B\) iff \(f\) ``doesn't care'' about the
      choice of \(A\).
      \begin{equation}
        \Pi {(x , y : A)} , f(x) \equiv f(y)
      \end{equation}
      Formally speaking, \(f\) needs to be ``coherently constant''
      \cite{krausGeneralUniversalProperty2015}, and \(B\) needs to be an
      \(n\)-type for some finite \(n\).
  \end{enumerate}
\end{rm-definition}
\subsection{Higher Homotopy Levels}
By hiding the position, we have essentially removed the ``decidable'' component
from split enumerability.
Our predicate now becomes general enough to work with non-sets: we will show
here that the circle is manifestly enumerable.
\begin{rm-theorem}
  The circle \(S^1\) is manifestly enumerable.
\end{rm-theorem}
\begin{proof}
  As the cover proof is a truncated proposition, we need only consider the point
  constructors, making this poof the same as the proof of split enumerability on
  \(\top\).
  \ExecuteMetaData[agda/Cardinality/Finite/ManifestEnumerable.tex]{circ-inst}
\end{proof}
\subsection{Surjections}
This predicates relation to surjectivity is much the same as split
enumerability's relation to \emph{split} surjectivity.
\begin{rm-lemma}
  A proof of manifest enumerability is equivalent to a surjection from a finite
  prefix of the natural numbers.
\end{rm-lemma}
\begin{proof}
  As with the other surjection proof (lemma~\ref{split-enum-surj}), this is
  simply a reassociation.
\end{proof}
\subsection{Relation to Split Enumerability}
\begin{rm-lemma}
  Any split enumerable type is also manifestly enumerable.
\end{rm-lemma}
\begin{proof}
  The proof carries over via truncation of the cover proof.
\end{proof}
\begin{rm-theorem} \label{manifest-enum-to-split}
  A manifestly enumerable type with decidable equality is split enumerable.
\end{rm-theorem}
\begin{proof}
  The support list stays the same between both enumerability proofs.

  For the cover proof, we need a way to get the contents out of a truncated
  value.
  We can do exactly that with the following lemma, called recompute:
  given a decision procedure for some type \(A\), and a truncation for the
  proposition \(A\), we can discount the possibility of \(A\) being false, and
  therefore extract the true decision.

  For the cover proof, our obligation now becomes constructing a decision for
  membership of the support list.
  This is straightforward given decidable equality.
\end{proof}
\subsection{Closure}
\begin{rm-lemma}
  Manifest enumerability is closed under dependent sum, disjoint union
  (non-dependent sum), and Cartesian product (non-dependent product).
\end{rm-lemma}
\begin{proof}
  For these three closures, the proofs on split enumerability consisted of a
  list manipulation followed by a proof that membership was preserved by the
  list manipulation.
  Because we separate these two concerns, the proofs carry over onto manifest
  enumerability: the support list manipulation stays the same, and the
  cover proofs are performed ``under'' the truncation.
\end{proof}
Notice that we do not have closure under functions: without decidability,
manifest enumerability is not closed under function arrows.
\section{Cardinal Finiteness} \label{cardinal}
For manifest enumerability, we removed the need for decidable equality: in these
next two finiteness predicates, we remove the need for a total order on the
underlying type.
\begin{rm-definition}[Cardinal Finiteness]
  A type \(A\) is cardinally finite, \(\mathcal{C}\), if it has a
  propositionally-truncated proof of bishop finiteness.
  \begin{equation}
    \mathcal{C}(A) = \lVert \mathcal{B}(A) \rVert
  \end{equation}
\end{rm-definition}
\subsection{Closure}
The closure proofs for cardinal finiteness are especially easy.
In contrast to manifest enumerability, under the propositional truncation we
have a full proof of bishop finiteness, meaning that all of the closure proofs
carry over.
\begin{rm-lemma}
  Cardinal finiteness is closed under dependent and non-dependent sums,
  products, and functions.
\end{rm-lemma}
\begin{proof}
  All closure functions can be lifted under propositional truncation.
  Therefore, cardinal finiteness has the same closure properties as manifest
  Bishop finiteness.
\end{proof}
\subsection{Decidable Equality}
\begin{rm-theorem} \label{cardinal-finite-discrete}
  Any cardinal-finite set has decidable equality.
\end{rm-theorem}
\begin{proof}
  We will use eliminator~\ref{elim-prop-prop} from definition~\ref{prop-trunc}.
  Manifest Bishop finiteness implies decidable equality already, so our task
  here is to prove that decidable equality itself is a proposition.

@@

  Notice that \(\mathbf{Fin}(n)\) and \(\mathbf{Fin}(m)\) are both equivalent to
  \(A\): we can join these proofs together, giving us the following:
  \begin{equation}
    \textbf{Fin}(n) \equiv \textbf{Fin}(m)
  \end{equation}
  Relying on the fact that \(\textbf{Fin}\) is injective
  (proposition~\ref{fin-inj}), we can derive that \(n\) and \(m\) are equal.
\end{proof}

\begin{rm-lemma} \label{fin-inj}
  \(\textbf{Fin}\) is injective.
\end{rm-lemma}
\begin{proof}
  Though this proof is surprisingly complex, it is a well-known puzzle in
  Martin-Löf type theory.
  Our proof does not differ significantly from standard approaches, so we will
  not detail it here.
\end{proof}
\subsection{Relation to Manifest Bishop Finiteness}
Cardinal finiteness tells us that there is an isomorphism between a type and
\(\mathbf{Fin}\); it just doesn't tell us \emph{which} isomorphism.
To take a simple example, \(\mathbf{Bool}\) has 2 possible isomorphisms with the
set \(\mathbf{Fin}(2)\): one where false maps to 0, and true to 1; and another
where false maps to 1 and true to 0.

@@

  Let \(\mathcal{X}\) be the proof of split countability on \(A\), and
  \(\mathcal{Y}\) be a function of the following type:
    \mathcal{Y} : \Pi {(x : A)} , \mathcal{E}!(U(x))
  Where \(U\) is a type family on \(A\).
  Our task is to provide the following:
    \mathcal{E}!\left(\Sigma {(x : A)} , U(x)\right)

  This final proof consists of the support stream, and the proof that the
  support stream covers the input.

  As mentioned, we will have to use a more sophisticated pairing function than
  the Cartesian product we used before.
  We instead will mirror the pattern in figure~\ref{cantor}.
  To greatly simplify the algorithm, we will produce an intermediate stream of
  lists which consists of the diagonals in the diagram.
  We then concatenate these streams into the final support stream.

>>>>

  We formalise our work in Cubical Agda
  \cite{vezzosiCubicalAgdaDependently2019}, and we implement a library for proof
  search (including combinators for level-polymorphic fully generic currying),
  and demonstrate how it can be used to both prove properties and synthesise
  full functions given desired properties.
\end{abstract}
\section{Introduction}
In this work, we will explore finite types in Cubical Type Theory
\cite{cohenCubicalTypeTheory2016}, and expose their relationship to infinite
types, species, and demonstrate their practical uses for proofs in dependently
typed programming languages.
\subsubsection{Strong Finiteness Predicates}
We will first explore the ``strong'' notions of finiteness (i.e. those at least
as strong as Kuratowski finiteness \cite{kuratowskiNotionEnsembleFini1920}),
with a special focus on cardinal finiteness (section~\ref{cardinal}), and
manifest enumerability (section~\ref{manifest-enumerability}), which is new, to
our knowledge.

\input{figures/finite-classification}
Figure~\ref{finite-classification} organises the predicates according to
their ``strength''; i.e. how much information they provide about a conforming
type.
For instance, a proof that some type \(A\) is manifestly Bishop finite (the
strongest of the notions, explored in section~\ref{manifest-bishop}) also tells
us that \(A\) is discrete (has decidable equality), and gives us a linear order
on the type.
A type that is Kuratowski finite (section~\ref{kuratowski}) has no such extra
features: indeed, we will see examples of Kuratowski finite types which are not
even sets, never mind discrete ones.

We will go through each of the predicates, proving how to weaken each (i.e. we
will provide a proof that every cardinally finite type is Kuratowski finite),
and how to strengthen them, given the required property.
In terms of figure~\ref{finite-classification}, this amounts to providing proofs
for each arrow.

We will---through the use of containers
\cite{abbottContainersConstructingStrictly2005}---formally prove the equivalence
these predicates have with the usual function relations i.e. we will show that a
proof of manifest enumerability is precisely equivalent to a surjection from a
finite prefix of the natural numbers.

For each predicate, we will also prove its closure over sums and products in
both dependent and non-dependent forms, if such a closure exists.
This will culminate in our main result for this section: the formal proof that
decidable Kuratowski finite sets form a topos.
More specifically, we show that cardinal finite sets form a topos, that
decidable Kuratowski finite sets are equivalent in strength to cardinal finite
sets, and carry the proof over.
This result relies on proofs on each of the other finiteness predicates.
\subsubsection{Species}
\todo[inline]{Redo this next paragraph}

\subsubsection{Infinite Types}
In section~\ref{infinite-cardinalities}, we will extend our study of finite
types to infinite but countable types.
We will see that the finiteness predicates are mirrored with countable
counterparts, and we will prove closure under the Kleene star and plus.

\subsubsection{Practical Uses of Finiteness}
Proofs of finiteness have well-known practical applications in
constructive mathematics \cite{firsovDependentlyTypedProgramming2015}.
In section~\ref{practical}, we build a library which exploits these uses in
Cubical Agda \cite{vezzosiCubicalAgdaDependently2019}, allowing automation of
complex proofs over finite types.
We frame this in terms of the principle of omniscience for finite types.
Thanks to the flexibility afforded to us by Cubical Type Theory, we are able to
go further than the usual examples of this kind of proof automation: as well as
proving properties about functions, we can synthesise functions whole-cloth from
their desired properties.
Through the unified interface for finite and countable types, we can
reuse the automation machinery for \emph{partial} proof search over infinite
search spaces.
Along the way, we extend the work in \cite{allaisGenericLevelPolymorphic2019} to
prove isomorphisms between the curried and uncurried forms of \(n\)-ary
dependent functions.
\section{Split Enumerability} \label{split-enumerable}
We will start with the simplest definition of finiteness: we say a set is
enumerable if there is a list of its elements which contains every element in
the set.
More formally:
\begin{rm-definition}[Split Enumerable Set]
  \begin{equation}
    \mathcal{E}!(A) = \Sigma {(\mathit{xs} : \textbf{List}(A))} , \Pi {(x : A)} , x \in xs
  \end{equation}
  We call the first component of this pair the ``support'' list, and the second
  component the ``cover'' proof.
\end{rm-definition}

@@

  It suffices now to prove the following:
  \begin{equation}
    \Pi {(x : A)} , \Pi {(\mathit{xs} : \textbf{List}(A))} , x \in \mathit{xs} \rightarrow x \inunique \text{uniques}(\mathit{xs})
  \end{equation}
  And from that we can generate our cover proof.
\end{proof}
Note that while manifest Bishop finiteness as split enumerability are equivalent
in ``strength'', the two types are not equivalent.
In particular, there are infinitely many inhabitants of \(\mathcal{E}!\), while
for a type \(A\) with \(n\) inhabitants, there are only \(n!\) inhabitants of
\(\mathcal{B}(A)\).
\subsection{Closure}
Proving equal strength of split enumerability and manifest Bishop finiteness
allows us to carry all of the previous proofs of closure over to manifest Bishop
finite sets (and vice-versa).
Missing from our previous proofs was a proof of closure of functions.
We remedy that here.
\begin{rm-theorem}
  Manifest bishop finiteness is closed over dependent functions
  (\(\prod\)-types).
  \begin{equation}
    \frac{
      \mathcal{B}(A) \; \; \; \Pi {(x : A)} , \mathcal{B}\left( U(x) \right)
    }{
      \mathcal{B}\left(\Pi {(x : A)} , U(x)\right)
    }
  \end{equation}
\end{rm-theorem}
\begin{proof}
  This proof is essentially the composition of two transport operations, made
  available to us via univalence.

@@

  We then prove an isomorphism between this representation and \(\Pi\)-types.
  \begin{equation}
    \mathbf{Tuple}(T, n) \iff \Pi {(x : \mathbf{Fin}(n))} , T(x)
  \end{equation}
  This allows us to transport our proof of finiteness on tuples to one on
  functions from \(\mathbf{Fin}\) (our second transport operation), proving our
  goal.
\end{proof}
\section{Manifest Enumerability} \label{manifest-enumerability}
A defining feature of split enumerability and Bishop finiteness is decidability:
both predicates imply a decidable equality function on the underlying type.
To find a predicate for finiteness which doesn't imply this, and therefore
works with types other than just sets, we \emph{truncate} the membership proof.
\begin{rm-definition}[Manifest Enumerability]
  \begin{equation}
    \mathcal{E}(A) = \Sigma {(\mathit{xs} : \mathbf{List}(A))} , \Pi {(x : A)} , \lVert x \in \mathit{xs} \rVert
  \end{equation}
\end{rm-definition}
\begin{rm-definition}[Propositional Truncation]a \label{prop-trunc}
  The type \(\lVert A \rVert\) on some type \(A\) is a propositionally truncated
  proof of \(A\).
  In other words, it is a proof that some \(A\) exists, but it does not tell you
  \emph{which} \(A\).

@@

  We will use and consume values of the type \(\lVert A \rVert\) in three main
  ways.
  \begin{enumerate}
    \item \label{elim-prop-prop} We can first eliminate from \(\lVert A \rVert\)
      into any type which is a proposition.
      In other words, given a function \(f : A \rightarrow B\), and a proof that
      \(B\) is a proposition, we can construct a function \(\lVert A \rVert
      \rightarrow B\).
    \item \label{elim-prop-monad} As a consequence of this first point, we can
      always eliminate into another propositionally truncated type.
      As a result, \(\lVert \cdot \rVert\) forms a Monad: for our purposes, this
      simply means tat we can work ``under'' a propositional truncation in an
      ergonomic way.
    \item \label{elim-prop-coh} We can eliminate from \(\lVert A \rVert\) with a
      function \(f : A \rightarrow B\) iff \(f\) ``doesn't care'' about the
      choice of \(A\).
      \begin{equation}
        \Pi {(x , y : A)} , f(x) \equiv f(y)
      \end{equation}
      Formally speaking, \(f\) needs to be ``coherently constant''
      \cite{krausGeneralUniversalProperty2015}, and \(B\) needs to be an
      \(n\)-type for some finite \(n\).
  \end{enumerate}
\end{rm-definition}

By hiding the position, we have essentially removed the ``decidable'' component
from split enumerability.
Our predicate now becomes general enough to work with non-sets: we will show
here that the circle is manifestly enumerable.
\begin{rm-theorem}
  The circle \(S^1\) is manifestly enumerable.
\end{rm-theorem}
\begin{proof}
  As the cover proof is a truncated proposition, we need only consider the point
  constructors, making this poof the same as the proof of split enumerability on
  \(\top\).
\end{proof}
\subsection{Surjections}
This predicates relation to surjectivity is much the same as split
enumerability's relation to \emph{split} surjectivity.
\begin{rm-lemma}
  A proof of manifest enumerability is equivalent to a surjection from a finite
  prefix of the natural numbers.
\end{rm-lemma}
\begin{proof}
  As with the other surjection proof (lemma~\ref{split-enum-surj}), this is
  simply a reassociation.
\end{proof}
\subsection{Relation to Split Enumerability}
\begin{rm-lemma}
  Any split enumerable type is also manifestly enumerable.
\end{rm-lemma}
\begin{proof}
  The proof carries over via truncation of the cover proof.
\end{proof}
\begin{rm-theorem} \label{manifest-enum-to-split}
  A manifestly enumerable type with decidable equality is split enumerable.
\end{rm-theorem}
\begin{proof}
  The support list stays the same between both enumerability proofs.

  For the cover proof, we first need the following function which searches a
  list for a particular element (given decidable equality on \(A\)).
  \begin{equation}
    \decin \; : \Pi(x : A) , \Pi(\mathit{xs} : \mathbf{List}(A)) , \mathbf{Dec}(x \in \mathit{xs})
  \end{equation}
  Where \(\mathbf{Dec}(A)\) is a decision on some type \(A\).

  We then need to convert a value of type \(\mathbf{Dec}(x \in \mathit{xs})\) to
  \(x \in \mathit{xs}\).
  We use the following to do that:
  \begin{equation}
    \text{recompute} : \mathbf{Dec}(A) \rightarrow \lVert A \rVert \rightarrow A
  \end{equation}
  A propositionally truncated value can be used to \emph{refute} its negation.
\end{proof}
\subsection{Closure}
\begin{rm-lemma}
  Manifest enumerability is closed under dependent sum, disjoint union
  (non-dependent sum), and Cartesian product (non-dependent product).
\end{rm-lemma}
\begin{proof}
  For these three closures, the proofs on split enumerability consisted of a
  list manipulation followed by a proof that membership was preserved by the
  list manipulation.
  Because we separate these two concerns, the proofs carry over onto manifest
  enumerability: the support list manipulation stays the same, and the
  cover proofs are performed ``under'' the truncation.
\end{proof}
Notice that we do not have closure under functions: without decidability,
manifest enumerability is not closed under function arrows.
\section{Cardinal Finiteness} \label{cardinal}
For manifest enumerability, we removed the need for decidable equality: in these
next two finiteness predicates, we remove the need for a total order on the
underlying type.
\begin{rm-definition}[Cardinal Finiteness]
  A type \(A\) is cardinally finite, \(\mathcal{C}\), if it has a
  propositionally-truncated proof of bishop finiteness.
  \begin{equation}
    \mathcal{C}(A) = \lVert \mathcal{B}(A) \rVert
  \end{equation}
\end{rm-definition}
\subsection{Closure}
The closure proofs for cardinal finiteness are especially easy.
In contrast to manifest enumerability, under the propositional truncation we
have a full proof of bishop finiteness, meaning that all of the closure proofs
carry over.
\begin{rm-lemma}
  Cardinal finiteness is closed under dependent and non-dependent sums,
  products, and functions.
\end{rm-lemma}
\begin{proof}
  All closure functions can be lifted under propositional truncation.
  Therefore, cardinal finiteness has the same closure properties as manifest
  Bishop finiteness.
\end{proof}
\subsection{Decidable Equality}
\begin{rm-theorem} \label{cardinal-finite-discrete}
  Any cardinal-finite set has decidable equality.
\end{rm-theorem}
\begin{proof}
  We will use eliminator~\ref{elim-prop-prop} from definition~\ref{prop-trunc}.
  Manifest Bishop finiteness implies decidable equality already, so our task
  here is to prove that decidable equality itself is a proposition.

@@

  Notice that \(\mathbf{Fin}(n)\) and \(\mathbf{Fin}(m)\) are both equivalent to
  \(A\): we can join these proofs together, giving us the following:
  \begin{equation}
    \textbf{Fin}(n) \equiv \textbf{Fin}(m)
  \end{equation}
  All that remains now is to prove that \(\mathbf{Fin}\) is injective.
  Though the proof is surprisingly complex, it is a well-known puzzle in
  Martin-Löf type theory.
  Our proof does not differ significantly from standard approaches, so we will
  not detail it here.
\end{proof}
\subsection{Relation to Manifest Bishop Finiteness}
Cardinal finiteness tells us that there is an isomorphism between a type and
\(\mathbf{Fin}\); it just doesn't tell us \emph{which} isomorphism.
To take a simple example, \(\mathbf{Bool}\) has 2 possible isomorphisms with the
set \(\mathbf{Fin}(2)\): one where false maps to 0, and true to 1; and another
where false maps to 1 and true to 0.

@@

  This final proof consists of the support stream, and the proof that the
  support stream covers the input.

  As mentioned, we will have to use a more sophisticated pairing function than
  the Cartesian product we used before.
  We instead will mirror the pattern in figure~\ref{cantor}.
  To greatly simplify the algorithm, we will produce an intermediate stream of
  lists which consists of the diagonals in the diagram.
  We then concatenate these streams into the final support stream.
\end{proof}
\begin{rm-lemma}
  Split countability is closed under non-dependent product and sum.
\end{rm-lemma}
\begin{proof}
  Follows from theorem~\ref{split-countability-sigma}.
\end{proof}
\subsubsection{Kleene Star}
While we lose some closures with the inclusion of infinite types, we gain some
others.
In particular, we have the Kleene star.
This means, in effect, that we have closure under lists.
\begin{rm-theorem}
  Split countability is closed under Kleene star.
  \begin{equation}
    \mathcal{E}!(A) \rightarrow \mathcal{E}!(\mathbf{List}(A))
  \end{equation}
\end{rm-theorem}
\begin{proof}
  As with the proof of closure under \(\sum\), our main task here is to figure
  out a way to arrange the indices such that 
\end{proof}
\subsection{Manifest Countability}
As we can quotient out the position information with finite types, so can we with
countable types.
\todo[inline]{fill in rest here}
\section{Practical Uses} \label{practical}
The theory of finite types in constructive mathematics, and in HoTT in
particular, is rich and interesting, as we hope we have demonstrated thus far.
As well as being theoretically interesting, however, the proofs and combinators
we have defined here are \emph{practically} useful.
\subsection{Omniscience}
In this section we are interested in restricted forms of the limited principle
of omniscience \cite{myhillErrettBishopFoundations1972}.
\begin{rm-definition}[Limited Principle of Omniscience]
  For any type \(A\) and predicate \(P\) on \(A\), the limited principle of
  omniscience is as follows:
  \begin{equation}
    \left( \Pi {(x : A)} , \mathbf{Dec}(P(x)) \right) \rightarrow \mathbf{Dec} \left( \Sigma {(x : A)} , P(x) \right)
  \end{equation}
  In other words, for any decidable predicate the existential quantification of
  that predicate is also decidable.
\end{rm-definition}
The limited principle of omniscience is non-constructive, but individual types
can themselves satisfy omniscience.
In particular, \emph{finite} types are omniscient.

There is also a universal form of omniscience, which we call exhaustibility.
\begin{rm-definition}[Exhaustibility]
  We say a type \(A\) is exhaustible if, for any decidable predicate \(P\) on
  \(A\), the universal quantification of the predicate is decidable.
  \begin{equation}
    \left( \Pi {(x : A)} , \mathbf{Dec}(P(x)) \right) \rightarrow \mathbf{Dec} \left( \Pi {(x : A)} , P(x) \right)
  \end{equation}
\end{rm-definition}

All of the finiteness predicates we have seen justify exhaustibility.
We will only prove it once, then, for the weakest:
\begin{rm-theorem}
  Kuratowski-finite types are exhaustible.
\end{rm-theorem}
\begin{proof}
  \todo[inline]{Proof}
\end{proof}

Omniscience is stronger than exhaustibility, as we can derive the latter from
the former:
\begin{rm-lemma} \label{omniscient-is-exhaustible}
  Any omniscient type is exhaustible.
\end{rm-lemma}
\begin{proof}
  For decidable propositions, we know the following:
  \begin{equation} \label{lem-conversion}
    \Pi (x : A) , P(x) \leftrightarrow \neg \Sigma (x : A) , \neg P(x) 
  \end{equation}
  To derive exhaustibility from omniscience, then, we run the predicate in its
  negated form, and then subsequently negate the result.
  The resulting decision over \(\neg \Sigma(x : A) , \neg P(x)\) can be
  converted into \(\Pi (x : A) , P(x)\).
\end{proof}
We cannot derive, however, that any exhaustible type is omniscient, as we do not
have the inverse of equation~\ref{lem-conversion}:
\begin{equation} \label{choice-conversion}
  \Sigma (x : A) , P(x) \leftrightarrow \neg \Pi (x : A) , \neg P(x)
\end{equation}
Such an equation would allow us to pick a representative element from any type,
which is therefore non-constructive.
In a sense, equation~\ref{lem-conversion} requires a form of LEM on the
proposition (i.e. requires it to be decidable), whereas
equation~\ref{choice-conversion} requires a form of choice.
Those finiteness predicates which are ordered do in fact give us this form of
choice, so the conversion is valid.
As such, all of the ordered finiteness predicates imply omniscience.
Again, we will prove it only for the weakest.
\begin{rm-theorem}
  Manifest enumerable types are omniscient.
\end{rm-theorem}
\begin{proof}
  \todo[inline]{Proof}
\end{proof}
Finally, we do have a form of omniscience for prop-valued predicates, as they do
not care about the chosen representative.
\begin{rm-theorem}
  Kuratowski finite types are omniscient about prop-valued predicates.
\end{rm-theorem}
\begin{proof}
  \todo[inline]{Proof}
\end{proof}
\subsection{Synthesising Pattern-Matching Proofs}
In particular, they can automate large proofs by analysing every possible case.
In \cite{firsovDependentlyTypedProgramming2015}, the \(\AgdaDatatype{Pauli}\)
group is used as an example.
\ExecuteMetaData[agda/Data/Pauli.tex]{def}
For this type, there are several practical tasks we hope to achieve with the
help of our finiteness combinators:
\begin{itemize}
  \item Define decidable equality on the type
  \item Define the group operation on the type
  \item Prove properties about the group operation
\end{itemize}
Unfortunately, the simple pattern-matching way to do many of these tasks is
prohibitively verbose.
As \(\AgdaDatatype{Pauli}\) has 4 constructors, \(n\)-ary functions on
\(\AgdaDatatype{Pauli}\) may require up to \(4^n\) cases.
A proof of decidable equality is one such function: it can be seen fully
worked-through in the appendix.

The alternative is to derive the things we need from \(\Enum!\) somehow.
First, then, we need an instance for \(\AgdaDatatype{Pauli}\):
\ExecuteMetaData[agda/Data/Pauli.tex]{inst}
From here we can already derive decidable equality, a function which requires 16
cases if implemented manually.

For proof search, the procedure is a well-known one in Agda
\cite{devrieseBrightSideType2011}: we ask for the result of a decision procedure
as an \emph{instance argument}, which will demand computation during
typechecking.
\ExecuteMetaData[agda/Cardinality/Finite/Search.tex]{auto}
And finally we can prove the property we wanted to on \(\AgdaDatatype{Pauli}\)
like so:
\ExecuteMetaData[agda/Data/Pauli.tex]{noninst-prf}
